---
title: Using the CEOdata package
author: Xavier Fernández-i-Marín
date: "`r format(Sys.time(), '%d/%m/%Y')` - Version `r packageVersion('CEOdata')`"
classoption: a4paper,justified
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the CEOdata package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(CEOdata)
```


CEOdata is a package that facilitates the incorporation of microdata (individual
responses) of public opinion polls in Catalonia into `R`, as performed by the "Centre
d'Estudis d'Opinió" (CEO, Opinion Studies Center). It has basically two main
functions with a separate purpose:

- **`CEOdata()`**: that provides the _data_ of the surveys directly into `R`.
- **`CEOmeta()`**: that allows the user to inspect the details of the available
  surveys (_metadata_) and to search for specific topics and get the survey
  details.

# `CEOdata()`: Get the survey data

The most comprehensive kind of data on catalan public opinion is the
"Barometer", that can be retrieved by default by the main function `CEOdata()`.

```{r message = FALSE}
library(CEOdata)
d <- CEOdata()
```

This provides a cleaned and merged version of all the available Barometers,
providing easy acces to the following number of responses and variables:
```{r}
dim(d)
```

```{r}
d
```

```{r}
names(d)[1:50]
```


## Specific barometers or time frames

`CEOdata()` allows you to select specific Barometers, by providing their internal register in the `reo` argument. 
The reo is the internal name that the CEO uses, and stands for "Registre
d'Estudis d'Opinió" (register of opinion studies), and is the main identificator
of the survey, also present in the table of meta data. Although many of them are
numbers, some have a number, a slash and another number, and therefore a
character vector must be passed. Only a single REO can be passed, as it is not
guaranteed that different data matrices share any column, and may refer to very
different topics.

For instance, to get only the data of the study with register "746" (corresponding to March 2013):

```{r}
d746 <- CEOdata(reo = "746")
d746
```

The function `CEOdata()` also allows to restrict the whole set of barometers based on specific time frames defined by a date with the arguments `date_start` and `date_end` using the YYYY-MM-DD format.

```{r}
b2019 <- CEOdata(date_start = "2019-01-01", date_end = "2019-12-31")
b2019
```


## Convenience variables

By default `CEOdata()` incorporates new variables to the original matrix. Variables that are created for convenience, such as the date of the survey. 
The CEO data not always provides a day of the month. 
In that case, 28 is used. These variables appear at the end of the dataset and can be distinguished from the original CEO variables because only the first letter is capitalized.

```{r}
tail(names(d))
```


# `CEOmeta()`: Access to the metadata of studies and surveys

The function `CEOmeta` allows to easily retrieve, search and restrict by time
the list of all the surveys produced by the CEO, which amounts to more than a
thousand as of end of 2021.

When called alone, the function downloads the latest version of the metadata
published by the center, in a transparent way, and caching its content so that
any subsequent calls in the same `R` session do not need to download it again.

```{r}
CEOmeta()
```

## Search for specific topics though keywords

The first relevant argument for `CEOmeta()` is `search`, which is a built-in
simple search engine that goes through the columns of the metadata containing
potential descriptive information () and returns the studies that contain such
keyword.

```{r}
CEOmeta(search = "Medi ambient")
```


It is also possible to pass more than one value to `search`, so that the search
includes them (either one of them OR any other).

```{r}
CEOmeta(search = c("Medi ambient", "Municipi"))
```

## Restrict by time

Metadata can be retrieved for a specific period of time, by using the arguments
`date_start` and `date_end`, also using the YYYY-MM-DD format. In this case the
dates that are taken into account are dates where the study gets into the
records, not the fieldwork dates.

```{r}
CEOmeta(date_start = "2019-01-01", date_end = "2019-12-31")
```


## Browse the CEO site

In addition, to the search engine and the restriction by time `CEOmeta()` also allows 
to automatically open the relevant URLs at the CEO domain that contain the details 
of the studies gathered with the function. This can be done setting the `browse`
argument to `TRUE`. However, there is a soft limitation of only 10 URLs to be
opened, unless the user forces to really open all of them (proceed with caution,
as this may open many tabs in your browser and leave your computer out of RAM in
some scenarios of RAM black holes, such as Chrome).

```{r, eval = FALSE}
CEOmeta(search = "Medi ambient a", browse = TRUE)
```

It is also possible to specify an alternative language, so the default catalan
pages are substituted by the automatic translations provided by Apertium (for
Occitan/Aranese) or Google Translate.

```{r, eval = FALSE}
CEOmeta(search = "Medi ambient a", browse = TRUE, browse_translate = "de")
```



# Extensions

Once you have retrieved the data of the surveys, it is trivial to work with
them. For instance, to get the overall number of males and females surveyed:

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```


```{r}
d |>
  count(SEXE)
```


Or to trace the proportion of females surveyed over time, across barometers:

```{r prop-females, fig.width = 8, fig.height = 4, fig.cap = 'Proportion of females in the different Barometers.'}
d |>
  group_by(BOP_NUM) |>
  summarize(propFemales = length(which(SEXE == "Dona")) / n()) |>
  ggplot(aes(x = BOP_NUM, y = propFemales, group = 1)) +
  geom_point() +
  geom_line() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  expand_limits(y = c(0, 1))
```


Alternatively, the metadata can alse be explored using the different topics
(tags, called "Descriptors") covered as reported by the CEO.

```{r tags, fig.width = 6, fig.height = 6, fig.cap = 'Prevalence of topics covered.'}
tags <- CEOmeta() |>
  separate_rows(Descriptors, sep = ";") |>
  mutate(tag = factor(stringr::str_trim(Descriptors))) |>
  select(REO, tag)

tags |>
  group_by(tag) |>
  count() |>
  filter(n > 5) |>
  ggplot(aes(x = n, y = reorder(tag, n))) +
    geom_point() +
    ylab("Topic")
```


Or by examining the time periods where there has been fieldwork in
quantitative studies, since 2018.

```{r fieldwork, fig.width = 8, fig.height = 10, fig.cap = 'Fieldwork periods.'}
CEOmeta() |>
  filter(`Dia inici treball de camp` > "2018-01-01") |>
  ggplot(aes(xmin = `Dia inici treball de camp`,
             xmax = `Dia final treball de camp`,
             y = reorder(REO, `Dia final treball de camp`))) +
  geom_linerange() +
  xlab("Date") + ylab("Surveys with fieldwork") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())
```



# Development and acknowledgement

The development of `CEOdata` (track changes, propose improvements, report bugs) can be followed at [github](https://github.com/ceopinio/CEOdata/).


If using the date and the package, please cite and acknowledge properly the CEO and the package, respectively.

<!-- # References -->


